---
title: "6_Relational_Data"
author: "BM"
date: "18/11/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

Relational data utilises verbs that work for a pair of tables.

Mutating joins: Adds new variables to one data frame from matching
observations in another. Filtering joins: Filter observations from one
data frame based on whether or not they match an observation in another
table. Set operations: treat observations as if they were set elements.

```{r, echo = FALSE}
library(tidyverse)
library(nycflights13)
```

```{r, echo = FALSE}
flights <- flights
```

There are five related data frames in this data set flights airlines
airports weather planes

You can express relationships as a drawing

# nycflights13

## Exercises 13.2.1

### 1.

To draw a route map of flights you would need flights and airports
joined to get the lattitude and longitude and destinations.

### 2.

The relationship line in the example should go from left to right oigin
to origin to faa. That is, weather, flights, airports.

### 3.

If the weather table contained the destination/location of weather we
would be able to link destination, destination, flights. That is,
weather, flights, airports.

### 4.

Examples of special days with less travel might be 4th July,
Thanksgiving, Christmas etc - Public Holidays. You could call it in
table by yy/mm//dd and name of the holiday and feeding it back into
flights. From there the number of flights on these days can be observed.

# Keys

Keys are the variables that link tables. They are variables or sets of
variables unique to each observation. Primary keys: uniquely identifies
a key in its own table. Foreign keys: uniquely identigies a key in a
foreign table.

Variables can also be both a foriegn, and a primary key. Due to the
importance of keys when joining it is best practice to check that keys
do actually uniquely identify observations. That is, does the variable
represent n\>1 observations. This can be done via a count (of variable)
and filtering \> 1.

```{r}
flights %>% 
  count(year, month, day, flight) %>% 
  filter(n > 1)
```

This shows that flight numbers are being recycled, so we can create a
surrogate number to identify flights. In this case the surrogate is row
number.

## Exercises 13.3.1

### 1.

```{r}
flights %>% 
  mutate(row_number = row_number()) %>% 
  select(row_number, everything())
```

This creates a surrogate key on row number.

### 2.

Identifying the keys in variours data sets

```{r, eval=FALSE}
install.packages("Lahman")
library(Lahman)
?Lahman

install.packages("babynames")
library(babynames)
?babynames

?diamonds
```

So for these datasets they links are; Lahman is playid Babynames is key
by name, year, sex Diamonds are \~50,000 observations. We could assign a
surrogate, or use x,y,z and hope for same specs. But they do repeat in
this dataset which is problematic.

# Mutating join

Recall that this is combining two tables and using a key to make one.'

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2
flights

flights2 %>% 
  select(-origin, - dest) %>% 
  left_join(airlines, by = "carrier")



```

Hence, this is called a mutating join because we are joining the tables
by introducing a new variable. Where mutating is the addition of a new
variable.

\#Understanding joins We have two basic tibbles below to illustrate how
joins work.

```{r}
x1 <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y1 <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

## Inner join

This is the simplest form of joins. These joins occur wherever keys are
equal. Example

```{r}
x1 %>% inner_join(y1, by = "key")
```

Notably this join drops off the third and fourth observations because
they do not occur in both tables. This can be problematic.

## Outer joins

Joins variables that appear in at least one of the variables. There are
three sub-types Left join: observations in x Right join: observations in
y Full join: keeps all observations in x and y. There are graphical
representations of these joins in the R4DS web book. The summary is left
and right joins will differentiate in regards to missing data in our
above examples. In a left join variable 3 in x will be kept, while in
right join it will keep variable 4 in y when joining the two data
frames. This needs to be considered when dealing with NAs. Some joins
will skip over missing data which is not necessarily the correct way to
treat them.

## Dulicate keys

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)

left_join(x, y, by = "key")
#> # A tibble: 4 x 3
#>     key val_x val_y
#>   <dbl> <chr> <chr>
#> 1     1 x1    y1   
#> 2     2 x2    y2   
#> 3     2 x3    y2   
#> 4     1 x4    y1
```

This causes 4 observations for 2 when joining which is problematic when
each identifier is meant to be unique. These mistakes are generally just
that.

## Defining key columns

```{r}
flights2 %>%  left_join(weather)



```

In this case we let the data decide what to join. This can be avoided by
having tidy and specific names. Alternatively, without changing the
tibble you could specify the join. An example of this being problematic
is trying to join flights to planes. Both data frames have a year
variable, however, one represents the year of manufacture while the
other is the year of the flight.

```{r}
flights2 %>% 
  left_join(airports, by = c("dest" = "faa"))
```

The above join shows joining the flights and airports tibbles via
specifying which avariables represent the same thing, but are not named
the same.

## Exercises 13.4.6

### 1.

Computing the average delay by destination. Below is a quick way to draw
a map of the US by state.

```{r airports, include = FALSE}
airports <- airports
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE))

airports_with_delay <- airports %>% 
  left_join(delays, by = c("faa" = "dest"))

airports_with_delay

airports_with_delay %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(x = lon, y = lat, colour = avg_delay, size = avg_delay)) +
    borders("state") +
    geom_point() +
    coord_quickmap()

```

Quickmap in this plot is to ensure that axis are scaled appropriately.

\#\#\#3. Previously we did delays by destination, instead
delays_by_plane

```{r}
delays_by_plane <- flights %>% 
  group_by(tailnum) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE))

delays_by_plane

delays_by_plane %>% 
  left_join(planes, by = "tailnum") %>% 
  ggplot(aes(x = year, y = avg_delay))+
  geom_jitter(alpha = 0.2)+
  geom_point()+
  geom_smooth()+
  coord_cartesian(xlim = c(1970, 2013), ylim = c(-80, 100))



```

Due to the way the geoms are layered in this graph call the smooth uses
the whole data set rather than than the zoomed in. The coord cartesian
is zooming in on the graph rather than truncating the data.

\#Filtering join Filtering joins match obersavtions in the same way as
mutation joins. But they affect the obs, rather than the variables
semi_join: keeps all observations in x that match y anti_join: drops all
observations in x that match y

Semi_joins are useful for matching summarise tables to original rows.

Below we find the top 10 nycflights13 destinations. This is done
differently to how we have in previous workshops.

```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)

top_dest

```

Now we form an example of an anti-join, where we drop some data

```{r}
flights %>% 
  anti_join(planes, by = "tailnum") %>% 
  count(tailnum, sort = TRUE)
 
```

## Exercises 13.5.1

### 1.

```{r}
flights %>% 
  left_join(planes, by = "tailnum") %>% 
  anti_join(planes, by = "tailnum") %>% 
  count(carrier, sort = TRUE)
```

This shows that AA and MQ carriers are responsible for most of the
missing data.

### 2.

Filtering flights to show planes that have flown at least 100 flights.
To do so you would filter the join by c(100, 100000000)

### 4.

Finding worst delays would group by date and summarise delays largest to
smallest which would give you a descending list of worst delays which
can be joined to the weather data. You could right join the filter, or
left join which will drop the data not relative to those dates.

```{r }
worst_delay <- flights %>% 
  group_by(year, month, day) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  mutate(two_day_delay = avg_delay + lag(avg_delay)) %>% 
  arrange(desc(two_day_delay)) %>% 
  head(1)

worst_weather_maybe <- weather %>% 
  semi_join(worst_delay)

```

There don't appear to be any major relations to determine this delay

# Join variables

Choosing your key is important. Test it before you start writing code
and manipulating the data. What might a good key now might not be great
if you go to add more data down the track. For example, using long and
alt to describe location might work now, but it doesn't describe a
unique location. lat lon would be better keys that would be
generalisable.

The join order is important too, for example joining three data sets
which are broader and less specific and decrease to fewer observations
with greater information. You should start with the broad data set and
left join to it. You will get NAs, but at least you won't be missing
observations.

# Set operations (not in the recording)

Intersect: only obs with x and y Union: unique obs with x and y setdiff:
obs with x and not y

Recall using intersect and union to join SA2 data at UE in qgis. Very
useful for merging data sets with different variables. These were
spatial data sets that could be determined with spatial boundaries.

\#Strings (second chapter in the recording) strings cna be manipulated
with stringr which is part of the tidyverse They are notated with ""

```{r}
string1 <- "This is a string"
class(string1)
some_numbers <- c(1,2,3,4)
class(some_numbers)
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

This code illustrates the use cases of '' and "". It is important to
close the "", or you will face issues. Sometimes star (\*) is used a
wildcard. To search for this you would need the speech marks. The escape
character in r is /. This is needed to search for quotes with wildcards.

Below shows how to search for different characters double quote = """
single quote = ''' back slash = "\\"" two back slashes = "\\\\"

There are non-english characters you can also short cut to. For example
mu is "\u00b5"

```{r}
x <- "\u00b5"
x
```

All stringr functions start with "str" which is very useful... there is
a cheat sheet for working with strings too which is very useful. Like
all the R cheat sheets.

To combine strings use c("x","y") which is a list of strings to be used
together. Recall that tidyverse uses the snake_case, not point.case when
calling for functions.

You can do useful things like linking strings to produce labels as below

```{r}
str_c("prefix-", 1:3, "-suffix")
```

This shows the prefixes and suffixes being added to an input variable
which can useful for generating names.

It is unlikely that I will have to use strings anytime soon, but perhaps
in the future I will. So the majority of this chapter is skipped over
for brevity's sake in the video and saving cognitive bandwidth. BUT, I
should cover this content at some point as I may need it at some point.

This leads into regular expressions, powerful tools than can get out of
hand.
